---
title: "PSY6422 Final Project"
author: "Maya Mardania"
date: "`r Sys.Date()`"
output: html_document
---

```{r markdown setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
# Prevents "Error in contrib.url():! trying to use CRAN without setting a mirror" message
options(repos = c(CRAN = "https://cloud.r-project.org"))

```

## Topic and Research Question

How do units of net housing supply compare across regions of England? Ajusting to regional population.

## Data
# Import

```{r library and data importing, include = FALSE, message = FALSE}

#readODS package needed to read the Net Additional Dwellings .ods dataset
if(!require(readODS)) install.packages("readODS")
library(readODS)
##readxl package needed to read the Regional Population .xlsx dataset
install.packages("readxl")
library(readxl)

#Libraries required for data wrangling
if(!require(dplyr)) install.packages("dplyr")
library(dplyr)
if(!require(tidyverse)) install.packages("tidyverse")
library(tidyverse)

#Required for plotting

#Adding data to a data frame
#Ensure data names match what I've named them below and are stored in a "data" file if you are downloading from sources yourself!
data_dwell <- read_ods("data/raw_netadditionaldwellings.ods", sheet = 5)
data_pop <- read_xlsx("data/raw_population.xlsx", sheet = 11)

#See our new imported datasets are what we need
View(data_dwell)
View(data_pop)
#Data needs to match by Year and Region, therefore need to ensure relevant wrangling is completed...

```

# Wrangle: Net Additional Dwellings data

```{r wrangle net dwellings data, include = FALSE, message = FALSE}
# Wrangle: Net Additional Dwellings data

#We are looking at regions of England, from 2011 to 2023 (as no pre-2011 population data was available unfortunately)
View(data_dwell)
#Looks like we need to remove some blank columns and rows, remove the England column, rename Year column data to just numbers, remove pre-2011 data, and convert the regions to long

#remove 1st 2 rows as they are irrelevant 
data_dwell <- data_dwell[-1:-2, ]

#make first row the headings for columns, then delete first row
colnames(data_dwell) <- data_dwell[1, ]
data_dwell <- data_dwell[-1, ]

#Check if that looks right
View(data_dwell) 

#Delete "England" column as we're just seeing regional differences
data_dwell <- data_dwell %>%
  select(-"England")

#select only Net data
#Backticks `` used to select labels with spaces in them
data_dwell <- data_dwell %>% filter(
  str_starts(`Components of net housing supply`, "Net additions")
  )

#can now delete this "Components of net housing supply" column as have used filters
data_dwell <- data_dwell %>%
  select(-`Components of net housing supply`)

#rename Year as first 4 digits: cannot convert to numeric otherwise
data_dwell$Year <- str_sub(data_dwell$Year, 1, 4)

#remove data pre-2011 to match the dataset with the population data we have
data_dwell <- data_dwell %>%
  filter(Year >= 2011)

#make long
data_dwell <- pivot_longer(data_dwell, 
             cols = -Year, #all columns except year as it's already long
             names_to = "Region",
             values_to = "Net_Additional_Dwellings")

#Quick check whether we have missing data
is.na(data_dwell)
#All values come back False, meaning no missing values

#Convert to numeric to allow plotting
data_dwell$Year <- as.numeric(data_dwell$Year)
data_dwell$Net_Additional_Dwellings <- as.numeric(data_dwell$Net_Additional_Dwellings)

#How does it look?
View(data_dwell)
#Numbers have shifted to the right indicating they are now numeric
#Looks good!

```

# Plot: Net Additional Dwellings data

```{r plot Net Additional Dwellings data, include = FALSE, message = FALSE}

#Quick view of our data

plot_dwell <- ggplot(data = data_dwell, mapping = aes(
  x = Year, y = Net_Additional_Dwellings, 
  group = Region, colour = Region)) + 
  geom_point() + 
  geom_line() +
  labs(title = "Net Dwellings by Region, 2011-2023", 
         x = "Year", 
         y = "Net Dwellings") +
    scale_x_continuous(breaks = seq(2011, 2023))

print(plot_dwell)

```

# Wrangle: Population data

```{r wrangle population data, include = FALSE, message = FALSE}

#Check imported data
View(data_pop)
#1st column, and first 6 rows can go. will also need to rename Year headings to only numeric, and convert to long

#remove first 6 rows, no data
data_pop <- data_pop %>% slice(-1:-6)

#remove first column, no data
data_pop <- data_pop[ ,-1]

#rename headers as the first row, then remove first row
colnames(data_pop) <- data_pop[1, ]
data_pop <- data_pop[-1, ]

#rename just 1st col to Region to match Dwellings data
colnames(data_pop)[colnames(data_pop) == "Name"] <- "Region"

#view only regional data
data_pop <- data_pop %>%
  filter(grepl("Region", Geography))

#remove Geography column as now not needed
data_pop <- data_pop[ ,-2]

#convert to long
data_pop <- data_pop %>%
  pivot_longer(
    cols = starts_with("Mid"),  # Select the year columns (2011, 2012, 2013)
    names_to = "Year",         # The new column for year
    values_to = "Population"   # The new column for population values
  )

# Remove 'Mid-' from the Year column
data_pop$Year <- gsub("Mid-", "", data_pop$Year)

#Change from UPPERCASE to LikeThis
library(tools)
data_pop$Region <- toTitleCase(tolower(data_pop$Region))

#Quick check whether we have missing data
is.na(data_pop)
#All values come back FALSE, meaning no missing values

#make numeric for plot
data_pop$Year <- as.numeric(data_pop$Year)
data_pop$Population <- as.numeric(data_pop$Population)

View(data_pop)
```

# Plot: Population

```{r population plot, include = FALSE, message = FALSE}

plot_pop <- ggplot(data_pop, aes(x = Year, y = Population, group = Region, colour = Region)) +
  geom_line() +
  geom_point() +
  labs(title = "Population by Region, 2011-2023", 
       x = "Year", 
       y = "Population") +
  scale_x_continuous(breaks = seq(2011, 2023))

plot_pop
```

# Wrangle: Merging datasets

```{r wrangling both datasets and merging, include = FALSE, message = FALSE}

#Will want to combine the datasets to form a new data frame so Region and Year remain as columns, and Net Dwellings and Population are new
#Names of Regions and Years must be the same for all data to merge correctly, lets make sure the names are matching

#Check if we have anything earlier than 2011
any(data_dwell$Year < 2011)
#False, great!
any(data_pop$Year < 2011)
#False, great!

unique(data_dwell$Region)
unique(data_pop$Region)
#Looks like there are some differences in Region names with East and Yorkshire... 
#I prefer "East of England" and "Yorkshire and The Humber" as names from the Dwellings df, so will rename these in the Population df
#rename "East" to just "East of England", and "Yorkshire and the Humber" to "... The Humber"
data_pop <- data_pop %>%
  mutate(
    `Region` = recode(`Region`, "East" = "East of England"),
    `Region` = recode(`Region`, "Yorkshire and the Humber" = "Yorkshire and The Humber")
  )

#We can now create a new data frame where new homes are adjusted by population for each region

#merging with Region and Year labels remaining
data_merge <- merge(data_dwell, data_pop, by.x = c("Region", "Year"), by.y = c("Region", "Year"))

#Check all 9 regions are here
unique(data_merge$Region)
#yep
#Check all years we're interested in are accounted for
unique(data_merge$Year)
#Yep
#ANy data lost in the merge, resulting in NA values?
is.na(data_merge)
#All FALSE, looks good to me!

View(data_merge)

#make new columns numeric for plotting
data_merge$Year <- as.numeric(data_merge$Year)
data_merge$Net_Additional_Dwellings <- as.numeric(data_merge$Net_Additional_Dwellings)
data_merge$Population <- as.numeric(data_merge$Population)

# Calculate the Population/Houses*1000 ratio, create new column for figures
data_merge$Dwell_By_Pop <- ((data_merge$`Net_Additional_Dwellings` / data_merge$`Population`)*1000)

#see the new column
View(data_merge)

```



# Final Plot for Assessment

```{r fig.width=11, fig.height=5}

#plot it all!
plot_merge <- ggplot(data_merge, aes(x = Year, y = Dwell_By_Pop, group = `Region`, colour = `Region`,
                                    text = paste(Region, "|", Year,
                                                 "<br>Population:", format(Population, big.mark = ","),
                                                 "<br>Completed Builds:", format(Net_Additional_Dwellings, big.mark = ","),
                                                 "<br><b>Homes completed per 1000 people:</b>", round(Dwell_By_Pop, 3)))) +
  geom_line() +
  geom_point() +
  labs(title = "New Homes Completed to Population by English Region, 2011-2023", 
       x = "Year",
       y = "New Homes completed per 1000 people") +
  scale_x_continuous(
    breaks = seq(2011, 2023)) +
  scale_y_continuous(
    breaks = seq(1, 5.5, by = 0.25),
      limits = c(1, 5.5))

plot_merge

library(plotly)

## Convert ggplot to plotly & ensure custom hover text is shown
plot_interactive_ratio <- ggplotly(plot_merge, tooltip = "text") %>%
  layout(legend = list(
    orientation = "h",  # Horizontal layout for the legend
      x = 0.5,            # Center the legend horizontally
      y = 1.1,            # Position above the plot area (adjust as needed)
      xanchor = "center", # Align the legend box horizontally to its center
      yanchor = "bottom"  # Align the legend box vertically to its bottom
    )
  )
  
## Show the interactive plot
plot_interactive_ratio
```


## Conclusions
## Limitations & Future ideas


## References

All data was extracted from the Office of National Statistics.

<b>Net Additional Dwellings data</b>: Table 118: annual net additional dwellings and components, England and the regions (ODS, 54KB) -<br> <a href="https://www.gov.uk/government/statistical-data-sets/live-tables-on-net-supply-of-housing#live-tables"target="_blank">https://www.gov.uk/government/statistical-data-sets/live-tables-on-net-supply-of-housing#live-tables</a> (Updated:
28 November 2024)

<b>Population estimates data</b>: Mid-2023: 2023 local authority boundaries edition of this dataset edition of this dataset (xlsx, 813.1KB) -<br> <a href="https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/estimatesofthepopulationforenglandandwales">https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/estimatesofthepopulationforenglandandwales</a> (Released: 15 July 2024)

